---
title: "Lab 4"
author: "Group 6"
date: "11/30/2020"
output: pdf_document
---

```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1: Computations with Metropolis Hastings

Target density : 
```{r}
f = function(x){
  res = x^5 * exp(-x)
  return(res)
}
```

## 1

Using Metropolis Hastings algorithm to generate samples from given distribution by using
proposal distribution as log normal LN(Xt; 1), take 1 as starting point.

```{r}
propose_lognorm_MCMC<-function(nstep,X0){
  vN<-1:nstep
  vX<-rep(X0,nstep);
  for (i in 2:nstep){
    X<-vX[i-1]
    Y<-rlnorm(1,mean=X,sd=1)
    u<-runif(1)
    a<-min(c(1,(f(Y)*dlnorm(X,meanlog = Y,sdlog = 1))/(f(X)*dlnorm(Y,meanlog = X,sdlog = 1))))
    if(u<=a){vX[i] = Y} else {vX[i] = X}   
  }
  return(vX)
}
set.seed(12345)
lognorm = propose_lognorm_MCMC(40000,1)

plot(1:40000,lognorm,pch=19,cex=0.3,type = "l",col="black",xlab="t",ylab="X(t)",main="")
```

The chain takes different amount of time to converge depending on the selected starting point , this period is called as burn-in period. In the above graph we can see that the chain did not converge. Hence there is no burn in period.

## 2

Using the chi square distribution $\chi^2([X_t+1])$ as a proposal distribution,
where $[x]$ is the floor function.

```{r}
propose_chi_MCMC<-function(nstep,X0){
  vN<-1:nstep
  vX<-rep(X0,nstep);
  for (i in 2:nstep){
    X<-vX[i-1]
    Y<-rchisq(1,df=floor(X+1))
    u<-runif(1)
    a<-min(c(1,(f(Y)*dchisq(X,floor(Y+1)))/(f(X)*dchisq(Y,floor(X+1)))))
    if(u<=a){vX[i] = Y} else {vX[i] = X}   
  }
  return(vX)
}
set.seed(12345)
chi = propose_chi_MCMC(10000,1)

plot(1:10000,chi,pch=19,cex=0.3,type = "l",col="black",xlab="t",ylab="X(t)",main="",ylim=c(0,20))
```
In the above chain we can observe that chain converges in the very begining , ie to say it follows a similar pattern fluctuating around a mean value. Convergence seem to have achieved near the starting point , so we can consider first few iterations as burn in period.

## 3

In conclusion we can say that chi squared is the better choice as a proposal distribution for the given target distribution, this can be justified by observing the time series plot for both proposal distributions, we could not find a proper convergence for log normal distribution , where as in sampling with chi squared convergence was very fast with a small burnin period.

## 4

```{r, warning=FALSE}
# reference : http://ugrad.stat.ubc.ca/R/library/coda/html/gelman.diag.html
library(coda)
set.seed(12345)
mcmc_data = mcmc.list()
for(i in 1:10){
  mcmc_data[[i]] = as.mcmc(propose_chi_MCMC(10000,i))
}

gelman.plot(mcmc_data)

gelman.diag(mcmc_data)
```
Two ways to estimate the variance of stationary distribution 
1. The mean of the emperical variance within each chain $W$
2. The emperical variance from all chain combined, this can be expressed as 
$$\sigma^2 = \frac{(n-1)W}{n} + \frac{B}{n}$$
where $n$ is number of iteration and $\frac{B}{n}$ is emperical between chain variance.

Assumption : Target distribution is normal.
A bayesian probaility interval can be constructed using t distribution with

$$\hat\mu = sample\space mean\space of\space all\space chains\space combined$$
variance $$\hat{V} = \sigma^2 + \frac{B}{mn}$$
degrees of freedom as  $$ d =\frac{2\hat{V}}{var(\hat{V})} $$
The convergence diagnostic iteself is given by :
$$ R = \sqrt{\frac{(d+3)\hat{V}}{(d+1)W}} $$
The value substantially above 1 indicates that the lack of convergence , we have obtained this factor as 1 confirming the convergence.

## 5

Estimate $\int_0^\infty xf(x) dx$

Step 1 : 
Mean of generated value in step 1 is given by 
$$\frac{1}{n} \sum_i x_i $$
where x is generated values with log normal as proposal density.

```{r}
mean(lognorm)
```
Step 2 : 
Here x is generated values with chi square as proposal density.

```{r}
mean(chi)
```

## 6

The probability density function of gamma distribustion is given by 

$$ f(x;\alpha , \beta) = \left[\frac{1}{\Gamma(\alpha)\beta^\alpha} \right] x^{\alpha-1}exp^\frac{-x}{\beta} $$

$$ 0 < x < \infty $$

Comparing given density function and gamma function , we can find alpha and beta as follows
$$\alpha - 1 = 5$$
ie $$ \alpha = 6 $$

$$ \frac{1}{\beta} = 1 $$ 
ie $$ \beta = 1$$

Mean or expected value for gamma distribution is given by $$\mu = \alpha\beta $$
ie 6*1 = 6

Our estimation in step 2 is `r mean(chi)` which is almost 6. This indicates that chi square is a good choice for proposal distribution for the given target distribution.

# Question 2: Gibbs sampling

## 1. Import the data to R and plot the dependence of Y on X. 
What kind of model is reasonable to use here?

```{r, echo=FALSE,message=FALSE,warning=FALSE,include=FALSE}
#--------------------------------------------------------------
#A2
#--------------------------------------------------------------
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
load("chemical.RData")
require(ggplot2)
data = data.frame("X" = X,
                  "Y" = Y)
x.2 <- lm(Y ~  poly(poly(X, 2)) ,
          data = data)
lg.x = lm(Y ~ log(X),
          data = data)

gg <- ggplot(data, aes(x = X, y = Y)) + 
  geom_point(color="purple")+
  ggtitle("concentration of a certain chemicals in water")+
  geom_line(aes(x = X, y = predict(x.2), colour = "x^2"))  +
  geom_line(aes(x = X, y = predict(lg.x), colour = "log (x)"))+  theme_minimal() +
  xlab("Day") + ylab("Concentration")

gg
## the model is noisy 

```
The model is noisy so a simple linear regression can not be used to describe the model. In the plot we have shown a second degree polynomial regression and log(x). Best on the plot, the second degree polynomial is the better choice.

2. A researcher has decided to use the following (random-walk) Bayesian model (n=number of observations, $\vec{\mu} = (\mu_1, ..., \mu_n)$ are unknown parameters):
$$ Y_i  = \mathcal{N}(\mu, \sigma = 0.2), i= 1,...,n$$
where the prior is
$$ p(\mu_1) = 1  $$
$$ p(\mu_{i+1} \mid \mu_i) = \mathcal{N}(\mu_i , 0.2) i = 1,...,n1  $$


*Present the formulae showing the likelihood $p(Y\ \mu)$ and the prior $p(\mu)$. Hint: a chain rule can be used here $p(\vec\mu) = p(\mu_1)p(\mu_2|\mu_1)...p(\mu_n|\mu_{n-1})$.*
the PDF of Gaussian Distribution is : 
$$\mathcal{N}(\mu,\,\sigma^{2}) = \frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(\vec{y_i}-\vec{\mu_i})^2}{2\sigma^2}}$$
and $\sigma^2 = 0.2$

### 2.2 Likelihood:
to obtain a formulae for likehood we should take product of this PDF:

$$ P(\vec{Y}|\vec{\mu}) = \prod_{i=1}^n \frac{1}{\sqrt{2 \pi 0.2^2}} exp(-\frac{(y_i-\mu_i)^2}{2*0.2^2})= 
\Big(\frac{1}{\sqrt{0.08 \pi}}\Big)^n exp \Big(-\frac{\sum_{i=1}^n(y_i-\mu_i)^2}{0.08} \Big)$$


Prior formulae:
$$ p(\vec{\mu}) = \prod_{i=1}^{n-1} \frac{1}{\sqrt{2 \pi* 0.2^2}} exp(-\frac{(\mu_{i+1}-\mu_i)^2}{2*0.2^2})$$
$$P(\vec{\mu}) = \bigg{(}\frac{1}{\sqrt{2\pi*0.2^2}}\bigg{)}^n \text{exp} \Bigg{[} -\frac{1}{2*0.2^2}\sum_{i = 1}^{n-1}(\vec{\mu_{i+1}}-\vec{\mu_i})^2 \Bigg{]}$$

### 2.3:
Use Bayes' Theorem to get the posterior up to a constant proportionality, and then find out the distributions of $(\mu_i|\mu_{-i}, Y)$, where $\mu_{-i}$i is a vector containing all $\mu$ values except of $\mu_i$.

From Bayes’ Theorem we know that:
$$Posterior\propto  likelihood*prior$$
hence,
$$P(\vec\mu|\vec{Y}) \propto \text{exp} \Bigg{[}  -\frac{1}{2\sigma^2}\sum_{i=1}^{n}({y_i}-{\mu_i})^2 \Bigg{]} \text{exp} \Bigg{[} -\frac{1}{2\sigma^2}\sum_{i = 1}^{n-1}({\mu_{i+1}}-{\mu_i})^2 \Bigg{]}$$

 if $n=1$ from Hint A:
 
$$P(\mu_1 | \vec\mu_{-1},\vec Y)  \propto \text{exp} \Bigg{[} -\frac{1}{2*0.2^2} \bigg{[} \mu_1 - \frac{y_1+\mu_2}{2} \bigg{]}^2 \Bigg{]} $$

 if $i\ne 1,n$ from Hint c:
 $$P(\mu_i | \vec\mu_{-i},\vec Y) \propto \text{exp} \Bigg{[} -\frac{3}{0.08} \bigg{[}  \mu_i - \frac{y_i+\mu_{i-1} + \mu_{i+1}}{3}  \bigg{]}^2 \Bigg{]} $$

 if $i=n$ from Hint B:

$$P(\mu_n | \vec\mu_{-n},\vec Y)  \propto \text{exp} \Bigg{[} -\frac{1}{0.04} \bigg{[} \mu_n - \frac{y_n+\mu_{n-1}}{2} \bigg{]}^2 \Bigg{]})$$

### 2.4

```{r, warning=FALSE, message=FALSE,echo=FALSE}
f.MCMC.Gibbs <- function (nsteps , mu0) {
   d <- length ( mu0 )
  mu <- matrix (0, nrow =nsteps , ncol = d)
  mu [1, ] <- mu0
  Y <- matrix (0, nrow =nsteps , ncol = d)
  Y[1, ] <- sapply (X=mu0 , FUN = function (mu){
    y <- rnorm (1, mean =mu , sd =0.2)
    return (y)
  })
  k = 1
  
  
  k = 1
  while (k < nsteps ) {
    
    #i=1
    mu[k+1, 1] <- exp (-1/ 0.04 *
                         (mu[k, 1] -(Y[k, 1]+ mu[k, 2]) /2) ^2) /( dnorm (Y[k, 1], mean =mu[k, 1],
                                                                                  sd =0.2) )
    # for i's between 1 to 50
    for (i in 2: (d -1) ) {
      mu[k+1, i] <- exp (-1/ ((2*0.04 )/3)*
                           (mu[k, i]-(Y[k, i]+ mu[k+1, i -1]+ mu[k, i +1]) /3) ^2) /( dnorm (Y[
        k, i], mean =mu[k, i], sd =0.2) )
    }
    #i=50
    mu[k, d] <- exp (-1/ 0.04 *(mu[k, d]-(Y[k, d]+ mu[k+1, d -1])
                                /2) ^2) /( dnorm (Y[k, d], mean =mu[k, d
    ], sd =0.2) )
    
    # update Y
    Y[k+1, ] <- sapply (X=mu[k+1, ], FUN = function (x){
      y <- rnorm (1, mean =x , sd =0.2)
      return (y)
    })
    
    k <- k+1
  }
  return (mu)
}
set.seed(147)
n <- length(Y)
x0 <- rep (0, n)

nsteps <- 1000
mu <- f.MCMC.Gibbs ( nsteps=nsteps , mu0 = x0 )
predicted_mu <- colMeans (mu)
df <- as.data.frame(data)
df$Prediction <- predicted_mu
gg2 <- ggplot(df, aes(x = X, y = Y, col = "Concentration")) + geom_point() +
  xlab("Day") + ylab("Concentration") +
  ggtitle("concentration of a certain chemicals in water") +
  geom_point(aes(x = X, y = Prediction, col = "Prediction")) + theme_minimal()
gg2
```
By using the formulae we calculated in 2.3 , we implemented a Gibbs Sampler. The algorithm ran for 1000 times. In 1000 time the Gibbs Sampler has not been able to capture the original data.

### 2.5:
trace plot of $\mu_n$ over its 1000 iterations.

```{r warning=FALSE,message=FALSE}
mu_samples <- as.data.frame(mu)
gg3 <- ggplot(mu_samples, aes(x = 1:nrow(mu_samples), y = mu_samples[,50])) + geom_line(color="lightblue") +
  ggtitle("Trace plot Mu_n")+ ylim(0,2.5)+theme_minimal()+xlab("Iteration") +ylab("Mu_n")
gg3
```
\newpage

# Appendix
```{r ref.label=knitr::all_labels(), echo = T, eval = F}