---
title: "Computational Statistics lab 1"
author: "Shwetha Vandagadde, Suhani Ariga and Hoda Fakharzadeh"
date: "11/4/2020"
output: 
    pdf_document:
      latex_engine: xelatex
header-includes:
    - \usepackage{caption}
    - \usepackage{float}
    - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

# Question 1: Be careful when comparing
```{r}

# ------------------------------------------------------------------------------
# A1
# ------------------------------------------------------------------------------

```
Two code snippets are mentioned in the question.
Snippet 1 : 
```{r  Question-1}
x1 = 1/3 
x2 = 1/4
if ( x1-x2 == 1/12) {
print ("Subtsraction is correct")
}else {
print ( " Substraction is wrong" )
}
```
Snippet 2 :
```{r}
x1 = 1
x2 = 1/2
if ( x1-x2 == 1/2) {
  print ("Subtsraction is correct")
}else {
  print ( " Substraction is wrong" )
}
```
Analysis :
 In case of first snippet , 1/3 and 1/12 results in a number with infinite decimal part, ie to say that these fractions has no finite representation and so R rounds it up. So the resulting value of the subtraction is not exactly equal to each other. So snippet 1 gives out messsage "Substraction is wrong"
 
But in case of snippet 2, 1/2 has a finite representation and hence snippet 2 gives out correct answer.

To go around the problem with snippet 1 , we can use all.equal since it tests for 'near equity' of two objects.
revised code for snippet 1

```{r}
x1 = 1/3 
x2 = 1/4
if (isTRUE(all.equal((x1 - x2),1/12))) {
  print ("Subtsraction is correct")
}else {
  print ( " Substraction is wrong" )
}

```
# Question 2: Derivative
```{r}

# ------------------------------------------------------------------------------
# A2
# ------------------------------------------------------------------------------

```
## Function to calculate derivative :

```{r}
f = function(x){x}

derivative_func = function(f=f,x){
  e = 10^-15
  f1 = (f(x + e) - f(x)) / e
  return(f1)
}
```
## Evaluating function at x = 1 and x = 100000

```{r}
x1 = derivative_func(f,1)
x1
```
```{r}
x100000 = derivative_func(f,100000)
x100000
```

## Results obtained and true values
As f(x) = x , true value is e/e ie 1.

For x = 1
e is not considered to be very small when x = 1, so f(1+e) results in 1.0000000000000011 which is slighlty greater than 1, after this gets substracted with f(1), we get numerator slightly greater than denominator and when this gets divided by e due to underflow in r , we get the answer slightly bigger than 1.

For x = 100000
When x is 100000 , value of e is very small , this results in numerator being 0 due to underflow error. Hence we get derivative for this function for a large x as 0


# Question 3: Variance
```{r}

# ------------------------------------------------------------------------------
# A3
# ------------------------------------------------------------------------------

```
## 1 My variance function

```{r}
myvar = function(x){
  n = length(x)
  var = (1/(1-n)) * (sum(x^2) - ((1/n) * (sum(x))^2))
  return(var)
}
```

## Generating x vector

```{r}
x = rnorm(10000, mean = 10^8, sd = 1)
```

## Plotting graph to compare the above variance function with var()

```{r}
n = length(x)
i = 1:10000
Yi = 0
for( j in 1:n){
  Yi[j] = myvar(x[1:j]) - var(x[1:j])
}
plot(x = i , y = Yi)

```
We can see there is difference between our variance function and var(). If there was no difference , we would have observed all the points to be at zero as there would be no difference. As the value of x increases , the summation of x^2 and x used in our formula will have overflow, hence the result is not accurate. This results in the difference of variances.

## Better implementation variance estimator

Trying to implement formula of sample variance , var = sum(x - xbar)^2 / n-1 to improve the results.

```{r}
better_var = function(x){
  len = length(x)
  x_bar = mean(x)
  v = (sum((x - x_bar)^2 ))/ (len-1)
  return(v)
}

n = length(x)
i = 1:10000
Yii = 0
j=0
for( j in 1:n){
  Yii[j] = better_var(x[1:j]) - var(x[1:j])
}

plot(x = i, y = Yii, ylim = c(-3,3))

```
This function gives better result as the difference plotted in the graph is concentrated at zero. This is because the x - xbar is done before summation is applied onto it , this tackles the overflow problem better than the function written before.


# Question 4: Binomial Coefficient
```{r Question-4}

# ------------------------------------------------------------------------------
# A4
# ------------------------------------------------------------------------------

```
A) prod(1:n)/(prod(1:k)∗prod(1:(n−k))) 
B) prod((k+1):n)/prod(1:(n−k))    
C) prod(((k+1):n)/(1:(n−k)))  


**1.** For the above R expressions below values of n and k does not work properly     
When k>n we will get answer -Inf/Inf instead of 0 for all expressions.

```{r}
n = 8
k = 14
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

When k=n we will get answer Inf instead of 1 for all expressions.

```{r}
n = 8
k = 8
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

When k<0 we will get answer -Inf/Inf instead of 0 for expression A.

```{r}
n = 10
k = -12
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

When k=0 we will get answer Inf instead of 1 for expression A.

```{r}
n = 10
k = 0
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

When n=k=0 we will get answer NaN instead of 1 for all expressions.

```{r}
n = 0
k = 0
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

**2.** 
## Comparing $A,B$ and $C$ for large values

```{r}
n = c(148, 148, 200, 320, 1500)
k = c(101, 100, 179, 191, 990)
i = 0
for(i in 1:5){
A = prod(1:n[i])/(prod(1:k[i])*prod(1:(n[i]-k[i])))
B = prod((k[i]+1):n[i])/prod(1:(n[i]-k[i]))
C = prod(((k[i]+1):n[i])/(1:(n[i]-k[i])))
D = choose(n[i], k[i])
cat("A expression    : ", A, fill = TRUE)    
cat("B expression    : ", B, fill = TRUE)    
cat("C expression    : ", C, fill = TRUE)
cat("choose function : ", D, fill = TRUE)
}
```

## Plotting the results from above expressions

```{r}  
A = c(1.048632e+39, 2.206498e+39, NaN)
D1 = c(1.048632e+39, 2.206498e+39, 1.383075e+28)
B = c(1.048632e+39, 2.206498e+39, 1.383075e+28, NaN)                                  
D2 = c(1.048632e+39, 2.206498e+39, 1.383075e+28, 2.300721e+92)                       
C = c(1.048632e+39, 2.206498e+39, 1.383075e+28, 2.300721e+92, Inf)
D3 = c(1.048632e+39, 2.206498e+39, 1.383075e+28, 2.300721e+92, Inf)
plot(A, D1, type = "o", col = "red")
plot(B, D2, type = "o", col = "blue")
plot(C, D3, type = "o", col = "orange")
```
Here, D1, D2 and D3 are the results from choose function for A, B and C expression respectively.
A expression throws overflow problem for n and k > 170 giving Inf/Inf = NaN. B expression throws overflow problem for n = 320 and k = 191 giving Inf/Inf = NaN. C expression gives same result as choose function(function to calculate binomial coefficient) and throws overflow problem for very large numbers. 


## Comparing $A,B$ and $C$ for small values

In the following plot we show the result for $A,B$ and $C$ when n = 173 and for k in range (1:170):
```{r Comparison_for_small_value}
f1 <- function(n,k){
  fact1 <- prod(1:n)/(prod(1:k)*prod(1:(n-k)))
  return(fact1)
}                                         # A expression
f2 <- function(n,k){
  fact2 <- prod((k+1):n)/prod(1:(n-k))    # B expression
  return(fact2)
  
}
f3 <- function(n,k){
  fact3 <- return(fact3)
}                                         #C expression


N = 170
K = 169
fact1 <- vector()
for(i in 1:K){
  fact1 <- append(fact1,f1(N,i))
}
plot(1:K,y = fact1,type = "l",col="blue",ylab = "bionomial coefficient", 
     xlab = "for k < 169")

fact2 <- vector()
for(i in 1:K){
  fact2 <- append(fact2,f1(N,i))
}
points(1:K,y = fact2,type = "p",col="green",cex=1,pch = 2)

fact3 <- vector()
for(i in 1:K){
  fact3 <- append(fact3,f1(N,i))
}
points(1:K,y = fact2,type = "p",cex=0.8,pch = 20,col="red")
legend("topright",legend=c("A","B","C"),
       col=c("blue","green" ,"red"), lty=c(1,NA,NA), 
       pch = c(NA,2,20),cex=0.8, text.font=4, box.lty=2,
       box.lwd=2, box.col="black",bg='lightyellow')

```
For small numbers, A, B and C expressions produces same set of results whereas C expression calculates binomial coefficient for large numbers compared to A and B expressions. 

**3.** For the below value of n and k, A expression gives Nan which is a overflow problem whereas B and C expressions produces the result. The A expression computes product from 1 to n (i.e.199) which exceeds the range of R. Hence it throws overflow problem first among 3 expressions.
```{r}
n = 199
k = 197
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

For the below value of n and k, A and B expressions gives NaN which is a overflow problem whereas C expression produces the result. The B expression computes product from k+1 to n and k is smaller. Hence it throws overflow problem.
```{r }
n = 199
k = 19
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

For the below value of n and k, A and B expressions gives Nan and C expression gives Inf which is a overflow problem. Both C expression and choose function(R built in function for binomial coefficient) gives same result. The C expression first computes the division operation and then prod() is used for the computed value. Here, the overflow problem occurs for very large numbers.
```{r}
n = 10037
k = 997
prod(1:n)/(prod(1:k)*prod(1:(n-k)))      # A expression
prod((k+1):n)/prod(1:(n-k))              # B expression
prod(((k+1):n)/(1:(n-k)))                # C expression
choose(n,k)
```

Hence we can conclude that C expression calculates binomial coefficient for  very large numbers compared to A and B expressions.

\newpage

# Appendix 

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```




